\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-16 docs/parco13/abstract.tex
Be more explicit on distributed memory

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% Clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (24 bytes per \pforest leaf plus 92 bytes per
forestclaw patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I hav added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).
\end{enumerate}

@@

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of ghost
cells surrounding each patch.  To this end, we define an interface
that allows distributed read-only access to the sequence of blocks,
the list of patches for each, and a lookup of neighbor patches (and
their relative orientation which is nontrivial between blocks).  The
exchange of ghost data and parallel partitioning is provided
transparently by \pforest.  Mesh modification directives, such as
local refinement and coarsening, are called from \forestclaw and
relayed to \pforest.

@@

\forestclaw uses iterators over all leafs, optionally restricted to a given
level.  Random access is possible and used when executing $O(1)$ time neighbor
lookups.  Looping over the patches in the order prescribed by the forest
% \comment{tree?}
% CB: I think forest is fine here.
and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse
\cite{BursteddeBurtscherGhattasEtAl09}.
% since the ghost patches are likely closeby with respect to the $z$-order.

>>>>

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% Clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf; \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I hav added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed.
\end{enumerate}

@@

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of ghost
cells surrounding each patch.  To this end, we define an interface
that allows read-only access to the sequence of blocks,
the list of patches for each, and a lookup of neighbor patches (and
their relative orientation which is nontrivial between blocks).
Just as \pforest, \forestclaw stores only the patches local to each
MPI rank.  The
exchange of ghost data and parallel partitioning is provided
transparently by \pforest.  Mesh modification directives, such as
local refinement and coarsening, are called from \forestclaw and
relayed to \pforest.

@@

\forestclaw uses iterators over all rank-local leafs, optionally restricted to
a given level.  Random access is possible and used when executing $O(1)$ time
neighbor lookups.  Looping over the patches in the order prescribed by the
forest
% \comment{tree?}
% CB: I think forest is fine here.
and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse
\cite{BursteddeBurtscherGhattasEtAl09}.
% since the ghost patches are likely closeby with respect to the $z$-order.
